\textbf{ Forecast time-scale determines which processes to model }

### 2019-04-22 doc/manuscript.tex
finished intro 1st draft

\title{\textbf{Integrating spatial and temporal patterns \\ in ecological forecasts }}

@@

%\textbf{\large{Keywords:}} Coexistence, competition, integral projection model, removal experiment, sagebrush steppe.

@@

We love ecological forecasting yada yada. Common approaches use spatial patterns, temporal patterns, and process-level understanding (i.e., mechanistic models). Since we will never know all the mechanisms, we will inevitably need to rely on patterns. But they have limitations.

The spatial approach: Correlate mean environment with the mean of an ecological state variable or rate across many sites. Examples: population distribution or abundance as a function of climate (SDM refs), primary production as a function of mean precipitation (Lauenroth refs).  Strengths: broad spatial scales, data are available and analyses are straightforward. Weaknesses: many well-recognized issues involving assumptions. A less well-recognized issue is that these models are not dynamic; they provide no information about how quickly the system will move from the current state to the predicted, future state. These patterns have developed over very long time spans, so they are unlikely to provide much information about dynamic changes on the order of years to decades. In addition, transient dynamics could prevent the system from ever reaching the predicted steady state (Urban ref?).

The temporal approach: Correlate interannual variation in an ecological response with interannual variation in the environment, typically at just one site (but see Kleinhesselink  and Adler 2018, others). Examples: population or vtial rate fluctuations as a function of weather, primary production as a function of precipitation (Lauenroth and Sala 1992). Strengths: focuses on temporal dynamics, can predict rates of change, easily validated. Weaknesses: typically limited spatial extent, ignores slower processes, such as evolutionary adaptation, or turnover in community composition, that could alter the underlying process at longer time scales.

Our goal is to capitalize on the complementary strengths of the spatial and temporal approaches. The idea is to take predictions from a spatial and a temporal model and combine them with a weighted average. The model weights can change over time: we expect the temporal model to dominate in the short-term, and the spatial model to dominate in the long-term. How rapidly that transition occurs depends on the rate of those ``slow" processes such as evolution and species turnover. We might learn something interesting by studying the rate at which the model weights change.

\section*{Eco-evo example}

>>>>

\title{\textbf{ Forecast time-scale determines which processes to model }}

@@

\textbf{\large{Keywords:}} dispersal, ecological forecasting, eco-evolutionary dynamics, global change, selection

@@

A recent flurry of synthesis papers in high impact journals attests to surging interest in ecological forecasting. Many of the papers 
argue that a commitment to ecological forecasting is necessary to guide the environmental policy and management
decisions that will become increasingly critical under global change (Clark 2001, Mouquet et al. 2017, Dietze et al. 2018).
A second argument is that rigorous testing of predictions will accelerate the process of basic research and discovery (Houlahan et al. 2017, Dietz 2017, Dietz et al. 2018).
Taken together, these papers, along with a new textbook (Dietze 2017) represent an emerging consensus that forecasting should be a priority for ecological research.

No such consensus is available for exactly how to tackle this challenge.
Ecologists rely on many different approaches and models for making predictions, and 
frequently argue about the virtues of process-based vs. phenomenological models (refs?).
We often justify investments in basic research by claiming that improved process-level understanding will improve predictions (e.g., Adler et al. 2012),
and it is easy to find examples of how extrapolation of linear relationships leads to poor predictions under novel conditions (refs?).
But in many cases simple phenomenological models make better predictions than complex process-based models with high parameter
uncertainty (refs? Tredennick et al. 2017?).

The pragmatic solution to these debates is clear: we should forecast using the models that make the best predictions.
Identifying these models requires rigorously testing predictions against observations. However, many common ecological predictions,
such as a species' geographic range in the year 2100, are virtually impossible to test (but see paleo hindcasting refs,Williams refs? Blois?), 
leading Dietze et al (2017) to call for greater emphasis on near-term forecasting. The assumption is that building 
our capacity to make short-term forecasts will ultimately improve our long-term forecasts.

Our goal is to consider a hypothesis that, in our view, has not been carefully considered by either the push for near-term forecasting or
ongoing debates about competing modeling approaches. We expect that short and long-term forecasts will require models
incorporating different kinds of processes and fit to different kinds of data sets. We demonstrate this idea using two simulation case studies, the first for an 
eco-evolutionary scenario, the second concerned with feedbacks between community composition and ecosystem functioning.  
Our analyses show that: 
1. For short-term forecasts, phenomenological time-series approaches are hard to beat.
2. Long-term forecasts require understanding slow processes such as evolutionary and ecological selection as well as dispersal.
3. Different kinds of data reflect the operation of different processes: longitudinal data capture autocorrelation and responses of current assemblages, while 
data spanning spatial gradients capture the long-term outcome of selection and dispersal.  As the time-scale of the forecast increases, the importance of these processes shifts.
4. A key challenge for future research is determining the rate at which slow processes first influence and ultimately dominate dynamics.

\section*{Modeling approach}

In both of our case studies, we simulate the effects of a hypothetical temperature increase on a simple system with known dynamics.
The ``truth'' is represented by a model that is fully mechanistic for at least one important process, but we assume that recovering 
this true, process-based model would be impossible in practice (we return to this point in the Discussion). We begin each simulation
under stationary temperature, allowing the system to equilibrate; we call this the baseline phase. We then increase temperature progressively
over a period of time, followed by a second period of stationary, now elevated, temperature. The objective is to forecast the response of the 
system to temperature increase based on data gathered during the baseline period.

We make two forecasts based on two different phenomenological models, each representing processes operating at different time scales. 
The first phenomenological model represents what we call a "spatial approach." We correlate the mean environment with the mean of an ecological state or rate 
 across many sites. This is the approach commonly used to predict population distribution or abundance as a function of climate (SDM refs) or
 primary production as a function of mean precipitation (Lauenroth refs). The approach is attractive because it addresses broad spatial scales relevant to 
 management, and because the necessary data are widely available and statistical analyses are straightforward. The approach also has well-known weaknesses.
Criticism often focuses on the assumption that current spatial patterns are in equilibrium with climate (refs?). A less well-recognized issue is that these models are not dynamic; 
they provide no information about how quickly the system will move from the current state to the predicted, future state. Because the patterns in the data have developed 
over long time spans, they are unlikely to provide much information about dynamic changes on the order of years to decades. Moreover, transient dynamics could 
prevent the system from ever reaching the predicted steady state (Urban ref?).

The second model represents what we call a ``temporal approach.'' We correlate interannual variation in an ecological response with interannual variation in the environment, 
typically at just one site. This approach is often used to study population or vital rate fluctuations as a function of weather, 
or primary production as a function of precipitation (Lauenroth and Sala 1992). The advantages of this approach include is focus on temporal dynamics, meaning it 
can predict rates of change, and validation of the resulting near-term predictions are straightforward. However, models built using this approach typically cover 
a limited spatial extent (but see Kleinhesselink  and Adler 2018, others), and ignore slower processes, such as evolutionary adaptation or turnover in community composition, 
that could alter the underlying process at longer time scales.

We compare forecasts from both models to the true, simulated dynamics to show that responses to environmental change reflect both kinds of processes, but at different time scales. To quantify this pattern, we combine forecasts from the spatial and temporal approaches with a weighted average and then show how the optimal model weights change over time: we expect the temporal model to best predict short-term dynamics, and the spatial model to best predict long-term dynamics. How rapidly that transition occurs depends on the rate of  ``slow" processes such as evolution and species turnover.

\section*{Eco-evolutionary example}
